Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-10 latex/main.tex

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{
Illustrative subgraphs in the ARG. Subgraphs around the XA-associated
recombination node (A); the recombination node associated with a group of
``nested'' Pango Xs (B); and the MRCA of the non-recombinants XN and XAU (C).
For more detailed visualizations of these subgraphs, see Document S2.
}
\label{fig:subgraphs}
\end{figure}

% This table is generated by the notebook tab_pango_x_events
\begin{table}
\caption{ARG events associated with Pango X lineages.
Each row corresponds to the oldest node assigned the specified
Pango lineage and the ancestor of all nodes assigned that label
(except XBB, XBB.1 and XM; see STAR methods). [MORE DETAILS]
\label{tab:pango_x_lineages}}
\begin{footnotesize}
\begin{tabular}{lrrrllrl}
\toprule
\multicolumn{5}{l}{Type I RE event: monophyletic for pango}\\
pango & avt & left & right & left p & right p & samp & nested \\
\midrule
XC & 37 & 26768 & 27390 & \textbf{AY.29} & \textbf{B.1.1.7} & 5 & {} \\
XBR & 22 & 22034 & 22190 & \textbf{BN.3.1} & BQ.1.25 & 1 & {} \\
XA & 16 & 20411 & 21765 & \textbf{B.1.177.18} & \textbf{B.1.1.7} & 39 & {} \\
XS & 11 & 9054 & 10449 & AY.103 & \textbf{BA.1.1} & 17 & {} \\
XL & 8 & 5925 & 8393 & \textbf{BA.1.17.2} & \textbf{BA.2} & 64 & {} \\
XBG & 8 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BA.5.2} & 25 & {} \\
XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
XQ & 7 & 4322 & 5386 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 55 & {BA.2:37,
XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
XAL:3} \\
XBF & 5 & 5184 & 9866 & \textbf{BA.5.2.1} & CJ.1 & 185 & {} \\
XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
XG & 4 & 5925 & 6513 & \textbf{BA.1.17} & \textbf{BA.2} & 3 & {} \\
XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
\midrule
\multicolumn{5}{l}{Type II RE event}\\
alias & avt & left & right & left p & right p & desc & by pango \\
\midrule
Xx & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 &
{BA.2:156, XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
XE/XH & 6 & 10448 & 11283 & \textbf{BA.1.17.2} & \textbf{BA.2} & 1156 &
{XE:1116, BA.2:37, XH:2, XAF:1} \\
XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {BA.2:4, XBH:2} \\
XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,
BF.3:2} \\ XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
XAF:1} \\
XJ & 5 & 13196 & 17410 & \textbf{BA.1.17.2} & \textbf{BA.2} & 85 & {XJ:68,
BA.2:17} \\
\bottomrule
\multicolumn{5}{l}{Events closely derived from RE}\\
pango & alias & path len & time & mutations & samples \\
\midrule
XR & XQ & 3 & 11 & 4 & 17 \\
XU & XQ & 6 & 103 & 4 & 1 \\
XAM & XQ & 6 & 43 & 3 & 21 \\
XAA & XQ & 6 & 33 & 3 & 17 \\
XAG & XQ & 6 & 55 & 4 & 6 \\
XAP & Xx & 4 & 67 & 1 & 20 \\
XZ & Xx & 4 & 48 & 1 & 48 \\
XAE & Xx & 2 & 47 & 7 & 9 \\
XAF & XAF & 5 & 140 & 8 & 1 \\
XAL & XM & 3 & 66 & 2 & 3 \\
\midrule
\multicolumn{5}{l}{Non recombinant}\\
pango & muts & parent & samples & nested \\
\midrule
XP & 1 & \textbf{BA.1.1} & 45 & {} \\
XN & 1 & \textbf{BA.2} & 120 & {BA.2:1} \\
XBQ & 1 & CJ.1 & 14 & {} \\
XBK & 1 & CJ.1 & 7 & {} \\
XBE & 3 & BA.5.2 & 65 & {} \\
XB & 2 & B.1 & 192 & {} \\
XAZ & 3 & \textbf{BA.5} & 133 & {} \\
XAV & 3 & BA.5.1.24 & 13 & {} \\
XAU & 1 & BA.2 & 8 & {} \\
XAS & 1 & BA.4 & 77 & {} \\
XAN & 5 & \textbf{BA.5.1} & 7 & {} \\
XAJ & 11 & \textbf{BA.2.12} & 18 & {} \\
\bottomrule
\end{tabular}
\end{footnotesize}
\end{table}

@@

To evaluate the concordance of the methods in
identifying the Pango lineages of recombinant parents and locating recombination breakpoints,
we focus on the Pango Xs associated with the Type I recombination events (Table~\ref{tab:pango_x_lineages}),
using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
For RecombinHunt and CovRecomb, we summarized the detection results for these Pango X lineages (if available)
reported in the original studies\cite{Alfonsi2024,Li2024CovRecomb} (STAR Methods).
For each method, we determined the Pango X lineages
for which the inferred parental lineages and breakpoint intervals
are concordant with those identified by the community (STAR Methods).
We found that the methods performed well overall,
with comparable concordance rates (Table~\ref{tab:method_concordance}),
indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.

@@

\subsubsection*{Identification of robust recombination events}
[TODO This section needs a full pass. I've merged in a section that was in the
supplement with an extremely terse section on counting supporting loci.]

@@

\subsubsection*{Comparison of mutational spectra}
To validate the mutational information contained in the sc2ts ARG,
we compared the all-site mutational spectra of major VOCs in the ARG,
a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024},
and another UShER phylogeny analyzed in a previous study\cite{Bloom2023}.
Mutational spectra are defined as the relative proportions of each possible type of nucleotide mutation.

First, we calculated the mutational spectra of Alpha, Delta, Omicron BA.1, BA.2, BA.4, and BA.5
in the sc2ts ARG, considering all the mutations above the sample nodes and internal nodes
which have Scorpio labels associated with the VOCs
(e.g., for Alpha, B.1.1.7-like and B.1.1.7-like+E484K).
These Scorpio labels were assigned to the aligned sequences of the nodes in the ARG using pangolin.
Note that $\sim$98\% of all the mutations in the ARG occur
above the nodes that have a Scorpio label associated with these VOCs
(excluding the nodes with unassigned Scorpio labels).

Next, we calculated the mutational spectra of the same VOCs
using the mutational counts from an UShER phylogeny analyzed in Bloom et al.\cite{Bloom2023},
which was built using Nextclade and has samples collected globally up to November, 2022.
We used the following mapping between the WHO labels and the Nextstrain clade labels:
Alpha, 20I; Detal, 21I and 21J; BA.1, 21K; BA.2, 21L; BA.4, 22A, and BA.5, 22B.
The mutation counts were downloaded from https://github.com/jbloomlab/SARS2-mut-spectrum/.

Lastly, we calculated the mutational spectra of the same VOCs
in the Viridian UShER phylogeny\cite{Hunt2024},
which was pruned down to the samples present in the sc2ts ARG.
We used the mutations above the samples based on their Scorpio labels,
which are available in the Viridian metadata provided by Hunt et al.\cite{Hunt2024}.

\subsubsection*{Comparison with recombination detection methods}
For RecombinHunt, we obtained two sets of detection results of the Pango X lineages
from the supplementary materials of Alfonso et al.\cite{Alfonsi2024}.
One set of results is based on mutational profiles calculated from a global GISAID dataset,
and the other set from a global NextStrain dataset.
The breakpoint intervals identified using RecombinHunt are defined
as the left and right recombination-informative sites.
We adjusted their coordinates by adding one to the right coordinate
so that the intervals are right-exclusive.

For CovRecomb, we summarized the detection results for the Pango X lineages as follows.
First, we obtained the "independent" recombination events
which correspond to the Pango X lineages (if available)
identified by Li et al.\cite{Li2024CovRecomb} (see Table S4 of the study).
When multiple events corresponded to a Pango X lineage,
we took the results of the event which had the highest number of epidemic recombinants.
For other Pango X lineages, we downloaded the full list of putative recombinants
identified in the study
(https://github.com/wuaipinglab/CovRecomb/blob/main/CovRecomb-Global-Version/putative\_recombinants/putative\%20recombinants.csv)
For each of these Pango Xs,
we took the most commonly observed combination of
the left parent, the right parent, and the breakpoint interval.
The breakpoint intervals are defined as the left and right recombination-informative sites.
We adjusted the breakpoint intervals by adding one to the right coordinate
so that they are right-exclusive.

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
% Referenced by the main Results.
We found that the ARG captures highly prevalent deletions in SARS-CoV-2 (Table~\ref{tab:major_dels}).
Li et al.\cite{Li2023indels} performed a large-scale analysis of indels
using a global GISAID dataset (sampled up to August, 2022).
We examined the major indels reported in Table S1 of that study
which have a minimum frequency of 1\% (sequence count divided by 9,149,680 filtered sequences).
Most of these indels (all deletions) are encoded in the ARG (Table~\ref{tab:indels})
and have high frequencies in the ARG (sequence count divided by the total number of samples),
except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
During site remapping, the 9-base deletion at 11283 was split up into
a 5-base deletion at 11283 (count of 339867) and a 4-base deletion at 11288 (count of 871782).

\begin{table}
\caption{
Major deletion events in the ARG.
These events are observed in at least 10,000 samples encoded in the ARG.
The counts and frequencies of these events from Li et al.\cite{Li2023indels} are also shown.
}
\label{tab:major_dels}
\begin{tabular}{rrcrcrrrr}
\centering
\toprule & \multicolumn{6}{c}{Sc2ts ARG} & \multicolumn{2}{c}{Li et al. (2023)} \\
\cmidrule(lr){1-7} \cmidrule(lr){8-9}
Start & Length & Region & Node & Pango & Count & Frequency (\%) & Count & Frequency (\%)\\
\midrule
28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 & 3853692 & 42.1183 \\
28248 & 6 & ORF8 & 220186 & B.1.617.2 & 1114669 & 44.9073 & 3026186 & 33.0742 \\
22029 & 6 & S / S1 - NTD & 200039 & B.1.617.2 & 1112410 & 44.8163 & 3010665 & 32.9046 \\
11288 & 4 & ORF1AB / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 & N/A & N/A \\
28362 & 9 & N & 1436802 & B.1.1.529 & 871774 & 35.1216 & 3939164 & 43.0525 \\
21633 & 9 & S / S1 - NTD & 822854 & BA.2 & 532072 & 21.4359 & 1941962 & 21.2244 \\
11292 & 5 & ORF1AB / NSP6 & 822854 & BA.2 & 531894 & 21.4287 & N/A & N/A \\
6513 & 3 & ORF1AB / NSP3 & 851246 & BA.1 & 340765 & 13.7286 & 1984683 & 21.6913 \\
11283 & 5 & ORF1AB / NSP6 & 851246 & BA.1 & 339867 & 13.6924 & 8360 & 0.0914 \\
21765 & 6 & S / S1 - NTD & 851246 & BA.1 & 339681 & 13.6849 & 3179887 & 34.7541 \\
21988 & 8 & S / S1 - NTD & 851246 & BA.1 & 339651 & 13.6837 & 590 & 0.0064 \\
22194 & 3 & S / S1 - NTD & 851246 & BA.1 & 312706 & 12.5982 & 1833050 & 20.0340 \\
28271 & 1 & non-coding & 86456 & B.1.1.7 & 290446 & 11.7014 & 3853692 & 42.1183 \\
11288 & 9 & ORF1AB / NSP6 & 86456 & B.1.1.7 & 290408 & 11.6998 & 3040576 & 33.2315 \\
21991 & 3 & S / S1 - NTD & 86456 & B.1.1.7 & 288831 & 11.6363 & 912944 & 9.9779 \\
21765 & 6 & S / S1 - NTD & 86456 & B.1.1.7 & 290438 & 11.7010 & 3179887 & 34.7541 \\
21765 & 6 & S / S1 - NTD & 1265302 & BA.4 & 190425 & 7.6718 & 3179887 & 34.7541 \\
28361 & 1 & N & 1246012 & BA.5.2.1 & 20822 & 0.8389 & 287 & 0.0031 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\caption{
Deletion events associated with the Alpha variant in the ARG.
These events occur above node 86456, which represents the origin of Alpha.
}
\label{tab:alpha_dels}
\centering
\begin{tabular}{rrrrr}
\toprule
Start & Length & Region & Count & Frequency (\%)\\
\midrule
28271 & 1 & non-coding & 290446 & 11.7014 \\
11288 & 9 & ORF1AB / NSP6 & 290408 & 11.6998 \\
21991 & 3 & S / S1 - NTD & 288831 & 11.6363 \\
21765 & 6 & S / S1 - NTD & 290438 & 11.7010 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\caption{
Highly recurrent deletions.
}
\label{tab:rec_dels}
\centering
\begin{tabular}{rrcrr}
\toprule
Start & Length & Region & Occurrences \\
\midrule
29781 & 1 & non-coding & 12713 \\
29762 & 1 & non-coding & 11356 \\
29779 & 1 & non-coding & 10925 \\
29769 & 1 & non-coding & 10783 \\
29774 & 1 & non-coding & 10707 \\
29700 & 1 & non-coding & 8543 \\
29614 & 1 & ORF10 & 5041 \\
29555 & 1 & non-coding & 4816 \\
29543 & 1 & non-coding & 4806 \\
29541 & 1 & non-coding & 4804 \\
203 & 1 & non-coding & 3602 \\
222 & 1 & non-coding & 3441 \\
22013 & 6 & S / S1 - NTD & 2111 \\
21991 & 3 & S / S1 - NTD & 2087 \\
28254 & 1 & ORF8 & 1967 \\
27556 & 64 & ORF7a & 1281 \\
27579 & 3 & ORF7a & 1014 \\
28093 & 1 & ORF8 & 898 \\
21765 & 6 & S / S1 - NTD & 478 \\
27532 & 1 & ORF7a & 403 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Analysis of Pango lineage ``origination'' events}
\label{sec:supplement_pango_lineages}
To analyze how well the ARG reflects the phylogenetic structure implicit in the
Pango lineage naming system, we considered the Pango assignments generated by
Pangolin on the alignments for each node in the ARG (STAR Methods). If the ARG
perfectly reflected the Pango lineage structure, each lineage would form a
clade descending from a single originating node with that label. We identified
putative origination nodes among those labeled with a given lineage as the node
with the maximum number of descendant samples and the earliest time (if there
is a tie).

@@

\begin{table} \centering
\caption{\label{tab:jackson}Comparison of recombination breakpoint
intervals and parent lineages for Groups A-D
reported by Jackson et al.\ with the corresponding
recombination events in the sc2ts ARG.
The second column gives the number of sequences in the group.
See the text for details of the groups and the sequences included.
limited to the samples considered by Jackson et al.
The breakpoint coordinates in Table 1 of Jackson et al.\ have been altered as follows: 
we subtract one to the left coordinates and add one to the right coordinates 
to correspond to the \texttt{tskit} definition of inheritance on either side of a breakpoint,
and add one to the right coordinate to make the intervals right-exclusive. }
\begin{tabular}{lllr@{--}lr@{+}l}
\toprule
Group        & Sequences & Method & \multicolumn{2}{c}{Interval}
    & \multicolumn{2}{r}{Parent lineages} \\
\midrule
A (XA)       & 2   & Jackson        &  21,254&21,767 & B.1.177&B.1.1.7 \\
             &     &\texttt{sc2ts} &  20,411&21,765 & B.1.177.18&B.1.1.7 \\
\cmidrule{3-7}
B            & 2   & Jackson        &  6,527&6,956 & B.1.36.28&B.1.1.7  \\
             &     &\texttt{sc2ts} &   6,529&6,954 & B.1.36.28&B.1.1.7  \\
\cmidrule{3-7}
C            & 3   &Jackson         &  24,913&28,653 &  B.1.1.7&B.1.221 \\
             &     & \texttt{sc2ts} &  25,997&27,972 &  B.1.1.7&B.1.221.1 \\
\cmidrule{3-7}
D            & 3   & Jackson        &  21,574&23,065 &  B.1.36.17&B.1.1.7 \\
             &     & \texttt{sc2ts} &  22,445&23,063 &  B.1.36.39&B.1.1.7 \\
\bottomrule
\end{tabular}
\end{table}

>>>>

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{
Illustrative subgraphs in the ARG.
(A) Subgraph around the recombination event associated with the XA.
The "Group A" sequences identified by Jackson et al.\cite{Jackson2021}) are attached to this event.
(B) Subgraph around the recombination event associated with a group of ``nested'' Pango Xs; and
(C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
For more detailed visualizations of these subgraphs, see Document S2.
}
\label{fig:subgraphs}
\end{figure}

% This table is generated by the notebook tab_pango_x_events
\begin{table}
\caption{
ARG events associated with Pango X lineages.
Each row corresponds to the oldest node assigned the specified
Pango lineage and the ancestor of all nodes assigned that label
(except XBB, XBB.1 and XM; see STAR methods). [MORE DETAILS]
}
\label{tab:pango_x_lineages}
\begin{footnotesize}
\begin{tabular}{lrrrllrl}
\toprule
\multicolumn{5}{l}{Type I RE event: monophyletic for pango}\\
pango & avt & left & right & left p & right p & samp & nested \\
\midrule
XC & 37 & 26768 & 27390 & \textbf{AY.29} & \textbf{B.1.1.7} & 5 & {} \\
XBR & 22 & 22034 & 22190 & \textbf{BN.3.1} & BQ.1.25 & 1 & {} \\
XA & 16 & 20411 & 21765 & \textbf{B.1.177.18} & \textbf{B.1.1.7} & 39 & {} \\
XS & 11 & 9054 & 10449 & AY.103 & \textbf{BA.1.1} & 17 & {} \\
XL & 8 & 5925 & 8393 & \textbf{BA.1.17.2} & \textbf{BA.2} & 64 & {} \\
XBG & 8 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BA.5.2} & 25 & {} \\
XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
XQ & 7 & 4322 & 5386 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 55 & {BA.2:37,
XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
XAL:3} \\
XBF & 5 & 5184 & 9866 & \textbf{BA.5.2.1} & CJ.1 & 185 & {} \\
XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
XG & 4 & 5925 & 6513 & \textbf{BA.1.17} & \textbf{BA.2} & 3 & {} \\
XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
\midrule
\multicolumn{5}{l}{Type II RE event}\\
alias & avt & left & right & left p & right p & desc & by pango \\
\midrule
Xx & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 &
{BA.2:156, XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
XE/XH & 6 & 10448 & 11283 & \textbf{BA.1.17.2} & \textbf{BA.2} & 1156 &
{XE:1116, BA.2:37, XH:2, XAF:1} \\
XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {BA.2:4, XBH:2} \\
XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,
BF.3:2} \\ XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
XAF:1} \\
XJ & 5 & 13196 & 17410 & \textbf{BA.1.17.2} & \textbf{BA.2} & 85 & {XJ:68,
BA.2:17} \\
\bottomrule
\multicolumn{5}{l}{Events closely derived from RE}\\
pango & alias & path len & time & mutations & samples \\
\midrule
XR & XQ & 3 & 11 & 4 & 17 \\
XU & XQ & 6 & 103 & 4 & 1 \\
XAM & XQ & 6 & 43 & 3 & 21 \\
XAA & XQ & 6 & 33 & 3 & 17 \\
XAG & XQ & 6 & 55 & 4 & 6 \\
XAP & Xx & 4 & 67 & 1 & 20 \\
XZ & Xx & 4 & 48 & 1 & 48 \\
XAE & Xx & 2 & 47 & 7 & 9 \\
XAF & XAF & 5 & 140 & 8 & 1 \\
XAL & XM & 3 & 66 & 2 & 3 \\
\midrule
\multicolumn{5}{l}{Non recombinant}\\
pango & muts & parent & samples & nested \\
\midrule
XP & 1 & \textbf{BA.1.1} & 45 & {} \\
XN & 1 & \textbf{BA.2} & 120 & {BA.2:1} \\
XBQ & 1 & CJ.1 & 14 & {} \\
XBK & 1 & CJ.1 & 7 & {} \\
XBE & 3 & BA.5.2 & 65 & {} \\
XB & 2 & B.1 & 192 & {} \\
XAZ & 3 & \textbf{BA.5} & 133 & {} \\
XAV & 3 & BA.5.1.24 & 13 & {} \\
XAU & 1 & BA.2 & 8 & {} \\
XAS & 1 & BA.4 & 77 & {} \\
XAN & 5 & \textbf{BA.5.1} & 7 & {} \\
XAJ & 11 & \textbf{BA.2.12} & 18 & {} \\
\bottomrule
\end{tabular}
\end{footnotesize}
\end{table}

@@

To evaluate the concordance of the methods in
identifying the Pango lineages of recombinant parents and locating recombination breakpoints,
we focus on the Pango X lineages associated with the Type I recombination events (Table~\ref{tab:pango_x_lineages}),
using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
For RecombinHunt and CovRecomb, we summarized the detection results for these Pango X lineages (if available)
reported in the original studies\cite{Alfonsi2024,Li2024CovRecomb} (STAR Methods).
For each method, we determined the Pango X lineages
for which the inferred parental lineages and breakpoint intervals
are concordant with those identified by the community (STAR Methods).
We found that the methods performed well overall,
with comparable concordance rates (Table~\ref{tab:method_concordance}),
indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.

@@

\subsubsection*{Identification of robust recombination events}

[TODO This section needs a full pass. I've merged in a section that was in the
supplement with an extremely terse section on counting supporting loci.]

@@

\subsubsection*{Comparison of mutational spectra}

To validate the mutational information contained in the sc2ts ARG,
we compared the all-site mutational spectra of major VOCs in the ARG,
a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024},
and another UShER phylogeny analyzed in a previous study\cite{Bloom2023}.
Mutational spectra are defined as the relative proportions of
each possible type of nucleotide change.

First, we calculated the mutational spectra of Alpha, Delta, Omicron BA.1, BA.2, BA.4, and BA.5
in the sc2ts ARG, considering mutations (ignoring indels) above the sample nodes and internal nodes
which have Scorpio labels associated with the VOCs
(e.g., for Alpha, B.1.1.7-like and B.1.1.7-like+E484K).
These Scorpio labels were assigned to the aligned sequences of the nodes in the ARG using pangolin.
Note that $\sim$98\% of all the mutations in the ARG occur
above the nodes that have a Scorpio label associated with these VOCs
(excluding the nodes with unassigned Scorpio labels).

Next, we calculated the mutational spectra of the same VOCs
using the mutational counts from an UShER phylogeny analyzed in Bloom et al.\cite{Bloom2023},
which was built using a Nextclade pipeline and
has samples collected worldwide up to November, 2022.
To obtain the WHO label corresponding to a Nextclade-defined clade, we used the following mapping:
20I, Alpha; 21I and 21J, Delta; 21K, BA.1; 21L, BA.2; 22A, BA.4; and 22B, BA.5.
The mutation counts were downloaded from https://github.com/jbloomlab/SARS2-mut-spectrum/.

Lastly, we calculated the mutational spectra of the same VOCs
in the Viridian UShER phylogeny\cite{Hunt2024},
which was pruned down to the samples present in the sc2ts ARG.
We used mutations above the sample nodes,
for which Scorpio labels are available
in the Viridian metadata provided by Hunt et al.\cite{Hunt2024}.

\subsubsection*{Comparison with recombination detection methods}

For RecombinHunt, we obtained two sets of detection results of the Pango X lineages
from the supplementary materials of Alfonso et al.\cite{Alfonsi2024}.
One set of results is based on mutational profiles calculated from a global GISAID dataset,
and the other set from a global NextStrain dataset.
The breakpoint intervals identified using RecombinHunt are defined
as the left and right recombination-informative sites.
We adjusted their coordinates by adding one to the right coordinate
so that the intervals are right-exclusive.

For CovRecomb, we summarized the detection results for the Pango X lineages as follows.
First, we obtained the "independent" recombination events
identified by Li et al.\cite{Li2024CovRecomb} (see Table S4 of the study)
which correspond to the Pango X lineages (if available).
When multiple events corresponded to a Pango X lineage,
we took the results of the event which had the highest number of epidemic recombinants.
For other Pango X lineages, we downloaded the full list of putative recombinants
identified in the study
(https://github.com/wuaipinglab/CovRecomb/blob/main/CovRecomb-Global-Version/putative\_recombinants/putative\%20recombinants.csv)
For each of these Pango Xs,
we took the most commonly observed combination of
the left parent, the right parent, and the breakpoint interval.
The breakpoint intervals are defined as the left and right recombination-informative sites.
We adjusted the breakpoint intervals by adding one to the right coordinate
so that they are right-exclusive.

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
% Referenced by the main Results.
The ARG captures highly prevalent deletions in SARS-CoV-2 (Table~\ref{tab:major_dels}).
Li et al.\cite{Li2023indels} performed a large-scale analysis of indels
using a global GISAID dataset (sampled up to August, 2022).
We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
which have a minimum frequency of 1\% (count of sequences divided by 9,149,680 sequences analyzed).
All of these indels are deletions and have high frequencies in the ARG
(sequence count divided by the total number of samples),
except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
During site remapping, the 9-base deletion at 11283 was split into
a 5-base deletion at 11283 (count of 339867) and a 4-base deletion at 11288 (count of 871782).

<TODO> Highly recurrent deletions (Table~\ref{tab:rec_dels}).

\begin{table}
\centering
\caption{
Major deletion events in the ARG.
These events are observed in more than 10,000 samples in the ARG.
}
\label{tab:major_dels}
\begin{tabular}{rrcrcrr}
\toprule
Start & Length & Region & Node & Pango & Count & Frequency (\%)\\
\midrule
28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 \\
28248 & 6 & ORF8 & 220186 & B.1.617.2 & 1114669 & 44.9073 \\
22029 & 6 & S / S1 - NTD & 200039 & B.1.617.2 & 1112410 & 44.8163 \\
11288 & 4 & ORF1AB / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 \\
28362 & 9 & N & 1436802 & B.1.1.529 & 871774 & 35.1216 \\
21633 & 9 & S / S1 - NTD & 822854 & BA.2 & 532072 & 21.4359 \\
11292 & 5 & ORF1AB / NSP6 & 822854 & BA.2 & 531894 & 21.4287 \\
6513 & 3 & ORF1AB / NSP3 & 851246 & BA.1 & 340765 & 13.7286 \\
11283 & 5 & ORF1AB / NSP6 & 851246 & BA.1 & 339867 & 13.6924 \\
21765 & 6 & S / S1 - NTD & 851246 & BA.1 & 339681 & 13.6849 \\
21988 & 8 & S / S1 - NTD & 851246 & BA.1 & 339651 & 13.6837 \\
22194 & 3 & S / S1 - NTD & 851246 & BA.1 & 312706 & 12.5982 \\
28271 & 1 & non-coding & 86456 & B.1.1.7 & 290446 & 11.7014 \\
11288 & 9 & ORF1AB / NSP6 & 86456 & B.1.1.7 & 290408 & 11.6998 \\
21991 & 3 & S / S1 - NTD & 86456 & B.1.1.7 & 288831 & 11.6363 \\
21765 & 6 & S / S1 - NTD & 86456 & B.1.1.7 & 290438 & 11.7010 \\
21765 & 6 & S / S1 - NTD & 1265302 & BA.4 & 190425 & 7.6718 \\
28361 & 1 & N & 1246012 & BA.5.2.1 & 20822 & 0.8389 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{
Deletion events associated with the Alpha variant in the ARG.
These events occur above node 86456, which represents the origin of Alpha.
}
\label{tab:alpha_dels}
\begin{tabular}{rrcrr}
\toprule
Start & Length & Region & Count & Frequency (\%) \\
\midrule
28271 & 1 & non-coding & 290446 & 11.7014 \\
11288 & 9 & ORF1AB / NSP6 & 290408 & 11.6998 \\
21991 & 3 & S / S1 - NTD & 288831 & 11.6363 \\
21765 & 6 & S / S1 - NTD & 290438 & 11.7010 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{Highly recurrent deletions.}
\label{tab:rec_dels}
\begin{tabular}{rrcr}
\toprule
Start & Length & Region & Occurrences \\
\midrule
29781 & 1 & non-coding & 12713 \\
29762 & 1 & non-coding & 11356 \\
29779 & 1 & non-coding & 10925 \\
29769 & 1 & non-coding & 10783 \\
29774 & 1 & non-coding & 10707 \\
29700 & 1 & non-coding & 8543 \\
29614 & 1 & ORF10 & 5041 \\
29555 & 1 & non-coding & 4816 \\
29543 & 1 & non-coding & 4806 \\
29541 & 1 & non-coding & 4804 \\
203 & 1 & non-coding & 3602 \\
222 & 1 & non-coding & 3441 \\
22013 & 6 & S / S1 - NTD & 2111 \\
21991 & 3 & S / S1 - NTD & 2087 \\
28254 & 1 & ORF8 & 1967 \\
27556 & 64 & ORF7a & 1281 \\
27579 & 3 & ORF7a & 1014 \\
28093 & 1 & ORF8 & 898 \\
21765 & 6 & S / S1 - NTD & 478 \\
27532 & 1 & ORF7a & 403 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Analysis of Pango lineage ``origination'' events}
\label{sec:supplement_pango_lineages}

To analyze how well the ARG reflects the phylogenetic structure implicit in the
Pango lineage naming system, we considered the Pango assignments generated by
Pangolin on the alignments for each node in the ARG (STAR Methods). If the ARG
perfectly reflected the Pango lineage structure, each lineage would form a
clade descending from a single originating node with that label. We identified
putative origination nodes among those labeled with a given lineage as the node
with the maximum number of descendant samples and the earliest time (if there
is a tie).

@@

\begin{table}
\centering
\caption{
Comparison of recombination breakpoint
intervals and parent lineages for Groups A-D
reported by Jackson et al.\ with the corresponding
recombination events in the sc2ts ARG.
The second column gives the number of sequences in the group.
See the text for details of the groups and the sequences included.
limited to the samples considered by Jackson et al.
The breakpoint coordinates in Table 1 of Jackson et al.\ have been altered as follows: 
we subtract one to the left coordinates and add one to the right coordinates 
to correspond to the \texttt{tskit} definition of inheritance on either side of a breakpoint,
and add one to the right coordinate to make the intervals right-exclusive.
}
\label{tab:jackson}
\begin{tabular}{lllr@{--}lr@{+}l}
\toprule
Group & Sequences & Method & \multicolumn{2}{c}{Breakpoint interval} & \multicolumn{2}{r}{Parent lineages} \\
\midrule
A (XA)       & 2   & Jackson        &  21,254&21,767 & B.1.177&B.1.1.7 \\
             &     &\texttt{sc2ts} &  20,411&21,765 & B.1.177.18&B.1.1.7 \\
\cmidrule{3-7}
B            & 2   & Jackson        &  6,527&6,956 & B.1.36.28&B.1.1.7  \\
             &     &\texttt{sc2ts} &   6,529&6,954 & B.1.36.28&B.1.1.7  \\
\cmidrule{3-7}
C            & 3   &Jackson         &  24,913&28,653 &  B.1.1.7&B.1.221 \\
             &     &\texttt{sc2ts} &  25,997&27,972 &  B.1.1.7&B.1.221.1 \\
\cmidrule{3-7}
D            & 3   & Jackson        &  21,574&23,065 &  B.1.36.17&B.1.1.7 \\
             &     &\texttt{sc2ts} &  22,445&23,063 &  B.1.36.39&B.1.1.7 \\
\bottomrule
\end{tabular}
\end{table}
